How does the writing system get informative? Differentiation or conservation?

### 2023-06-14 manuscript/wp4.tex
ms

Is it possible, then, that heterography may be selected for in the cultural evolution of orthography? And if so, what might be the mechanism through which such selection could operate? \textcite[pp.~325--326]{Berg:2021} sketch two models of the emergence of heterography (see Fig.~\ref{}). The first model, \textit{the differentiation model}, explains heterography as the result of the creation of a new orthographic form to differentiate one meaning from another. Imagine, for example, that English had chosen to adopt the French spelling <banque> for banks of the financial kind to differentiate them from banks of the river kind, resulting in a state of heterography: /baŋk/–/baŋk/ in speech, but <bank>–<banque> in writing. The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two spoken forms merge and become homophonous, but their heterographic spellings are conserved by the orthography. For example, the words \textit{night} and \textit{knight}, which are heterographic homophones in modern English, were previously pronounced differently from each other (in Old and Middle English, /kn/ was a permitted consonant cluster, as it still is in modern Dutch and German).

>>>>

Is it possible, then, that heterography may be selected for in the cultural evolution of orthography? And if so, what might be the mechanism through which such selection could operate? \textcite[pp.~325--326]{Berg:2021} sketch two models of the emergence of heterography (see Fig.~\ref{}). The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic form to differentiate between homophones. Imagine, for example, that English had chosen to adopt the French spelling <banque> for banks of the financial kind to differentiate them from banks of the river kind, resulting in a state of heterography: /baŋk/–/baŋk/ in speech, but <bank>–<banque> in writing. The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two spoken forms merge and become homophonous, but their heterographic spellings are conserved in the orthography. For example, the words \textit{night} and \textit{knight}---heterographic homophones in modern English---were previously pronounced differently from each other (in Old and Middle English, /kn/ was a permissible consonant cluster, as it still is in modern Dutch and German).

byte bite

bizarre bazaar

bombe

cache

<cheque>, which the OED describes as a ``a differentiated spelling of \textit{check}.''

Tindr

Innovation

French spellings
double letters inn
add -e

Processes

Sound change
dialect merger (bury is the spelling from one dialect with the pronunciation from another dialect)
Borrowing (English tends to retain spelling from the source language)
Printers innovations
- Etymological spellings (doubt)
- Deliberate design of heterographs
- Morphological spellings
